tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-13 paper.tex

The first step of simulating data for evaluating GWAS methods,
generating realistic genetic variation datasets of millions
of individuals, is therefore at a high-level of sophistication
and more than sufficient to support ongoing research.
The second step, of generating the actual quantiative traits,
is currently lagging far behind.
Although there are multiple packages to simulate genotypic data
concurrently with quantitative traits~\citep{gaynor2021,haller2023},
and to simulate quantitative traits from a given set of genetic
sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
handle the millions of individuals now required.
[TODO discuss bespoke stuff like Martin et al]
In particular, there is no software package available to simulate quantitative
traits directly from the efficient ARG representation used by
modern population genetic simulators.

>>>>

The first step of simulating data for evaluating GWAS methods,
generating realistic genetic variation datasets of millions
of individuals, is therefore at a high-level of sophistication
and seems sufficient to support ongoing research.
The second step, of generating the actual quantiative traits,
is currently lagging far behind.
Although there are multiple packages to simulate genotypic data
concurrently with quantitative traits~\citep{gaynor2021,haller2023},
and to simulate quantitative traits from a given set of genetic
sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
handle the millions of individuals now required.
% TODO this needs another pass. We do need to address the Martin et
% al thing up front though, I think.
\citet{martin2017} simulated quantitative traits for
600,000 humans as part of their analysis of the portability
of polygenic risk scores across populations. This was
done by using custom scripts to generate traits for these
large ARGs output by the \texttt{msprime}
simulator~\cite{kelleher2016efficient}.
While the code provided by Martin et al.\ was sufficient
for their analysis, it is tightly coupled to the details of their study
and not documented or designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and therefore there
is a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from the efficient ARG representation used by
modern population genetic simulators.
